---
title: "Belgium AIDS epidemic: solutions"
author: "Gavin Simpson"
date: "10/10/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Description

The data set `belgium.csv` contains the number of cases (`cases`) of AIDS recorded between 1981 and 1993 (`year`) in Belgium

## Packages

Load the following packages

```{r load-packages}
pkgs <- c("readr", "dplyr", "ggplot2", "here", "mgcv", "gratia")
vapply(pkgs, library, logical(1L), character.only = TRUE, logical.return = TRUE)
```

## Data

The data are contained in `data/belgium-aids.csv`, and consist of three variables

1. `year`, the year of each observation
2. `cases`, the number of AIDS cases per year

Load the data into R

```{r load-data}
belgium <- read_csv(here("data", "belgium-aids.csv"), col_types = "dd")
belgium
```

## Data wrangling

Create a variables `t` (for time) in the `belgium` data frame, which contains the time point of observation by subtracting `1980` from each `year`:

```{r wrangling}
belgium <- belgium %>%
  mutate(t = year - 1980)
```

## Task

1. Produce a plot of the number of AIDS cases over time

```{r}
belgium %>%
  ggplot(aes(x = year, y = cases)) +
    geom_point()
```

2. Fit a poisson GLM (log link) using the `glm()` to model the number of AIDS cases. Use `t` as the predictor variable

```{r}
m1 <- glm(cases ~ t, data = belgium, family = poisson(link = "log"))
```

3. Produce the model summary using `summary()`

```{r}
summary(m1)
```

4. Calculate a quick goodness of fit test for the model using the information provided on the residual deviance

```{r}
pchisq(deviance(m1), df.residual(m1), lower.tail = FALSE)
```

5. Plot the model diagnostics

```{r}
layout(matrix(1:4, ncol = 2, byrow = TRUE))
plot(m1)
layout(1)
```

6. Add fitted values of the model as a line on to the plot you created earlier. Add a confidence interval to the fitted value line.

```{r predict-m-1}
pdat <- with(belgium, tibble(t = seq(min(t), max(t),
                                           length = 100)))
## want year for each t; remember to add 1980
pdat <- pdat %>%
  mutate(year = 1980 + t)
pred <- predict(m1, newdata = pdat, type = "link", se.fit = TRUE)
ilink <- family(m1)$linkinv
pdat <- pdat %>%
  bind_cols(data.frame(pred)) %>%
  mutate(fitted = ilink(fit),
         upper = ilink(fit + (1.96 * se.fit)),
         lower = ilink(fit - (1.96 * se.fit)))

belgium %>%
ggplot(aes(x = year, y = cases)) +
  geom_ribbon(data = pdat,
              aes(ymin = lower, ymax = upper, x = year),
              alpha = 0.2, inherit.aes = FALSE) +
  geom_point() +
  geom_line(data = pdat, aes(y = fitted, x = year)) +
  labs(x = "Creatinine kinase level", y = "Proportion Heart Attack")
```

7. Looking at the model diagnostics and the plot of the fitted values versus the data, identify a problem with this model. Describe the problem in a few words.

8. Modify your model to account for the identified problem

    Hint: use ploynomials of `t`

```{r}
m2 <- glm(cases ~ poly(t, 2), data = belgium, family = poisson(link = "log"))

## or
m2a <- update(m1, . ~ . - t + poly(t, 2))
```

9. Look at the model summary, model diagnostic plots, and perform the goodness of fit test on the residual deviance. Are there any problems with this model?

```{r}
# model summary
summary(m2)
# quick GOF test
pchisq(deviance(m2), df.residual(m2), lower.tail = FALSE)
# model diagnostics
layout(matrix(1:4, ncol = 2, byrow = TRUE))
plot(m2)
layout(1)

# plots
pred <- predict(m2, newdata = pdat, type = "link", se.fit = TRUE)
ilink <- family(m2)$linkinv
pdat2 <- pdat %>%
  select(-c(fit, se.fit, residual.scale)) %>% # remove m1 link scale fit & std err
  bind_cols(data.frame(pred)) %>%
  mutate(fitted = ilink(fit),
         upper = ilink(fit + (1.96 * se.fit)),
         lower = ilink(fit - (1.96 * se.fit))) %>%
  bind_rows(pdat) %>%
  mutate(model = rep(c("t^2","t"), each = 100))

belgium %>%
ggplot(aes(x = year, y = cases)) +
  geom_ribbon(data = pdat2,
              aes(ymin = lower, ymax = upper, x = year, fill = model),
              alpha = 0.2, inherit.aes = FALSE) +
  geom_point() +
  geom_line(data = pdat2, aes(y = fitted, x = year, colour = model)) +
  labs(x = "Creatinine kinase level", y = "Proportion Heart Attack") +
  scale_fill_discrete(labels = scales::parse_format()) +
  scale_colour_discrete(labels = scales::parse_format())
```

10. Compare the two models using AIC and a generalized likelihood ratio test (GLRT). Which model is to be preferred?

    Hint: the `anova()` function is used to do the GLRT

```{r}
anova(m1, m2, test = "Chisq")

anova(m1, m2, test = "LRT")

pchisq(71.446, df = 1, lower.tail = FALSE)

deviance(m1)

deviance(m2)

deviance(m1) - deviance(m2)
```

## GAM

```{r}
# fit models using ML with gam() to be sure we get exactly the same loglik
m_gam0 <- gam(cases ~ poly(t, 2), data = belgium,
              family = poisson(link = "log"), method = "ML")
m_gam1 <- gam(cases ~ s(t), data = belgium, family = poisson(link = "log"),
              method = "ML")

# model summary for the GAM
summary(m_gam1)

# is GAM an improvement over the GLM? Yes, but no
anova(m_gam0, m_gam1, test = "Chisq")
AIC(m_gam0, m_gam1)
model_edf(m_gam0, m_gam1)
## differences in reported EDFs due to smoothness selection correction

## refit to use REML
m_gam2 <- update(m_gam1, . ~ ., method = "REML")

## plot the estimated smooth
draw(m_gam2, residuals = TRUE)

new_df <- tibble(year = seq(1981, 1993, length = 100)) %>%
  mutate(t = year - 1980)

pred_gam <- fitted_values(m_gam2, data = new_df, scale = "response")

plt <- pred_gam %>%
  ggplot(aes(x = year, y = fitted)) +
    geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) +
    geom_line() +
    geom_point(data = belgium, aes(x = year, y = cases))
plt

# compare with the GLM fits
plt + geom_line(data = pdat2, aes(colour = model))
```
